---
title: "CSC 308 Final Project — Individual R Portfolio"
author: "Ezaria Alexander"
date: "April 19, 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4.2)
```

# 1. Project Overview

This portfolio documents my individual contribution to the CSC 308 team project
*Exploratory Analysis of Spotify Audio Features Across Music Genres Using R.*
The team dataset was `genres_v2.csv` (Kaggle: mrmorj, 2020), sourced from the
Spotify Web API. The team produced six core visualizations — three distribution
histograms (danceability, energy, tempo), a scatter plot of energy vs. loudness,
a boxplot of danceability across the top 10 genres, and a correlation matrix of
all numeric audio features — and interpreted them in the context of four
research questions concerning feature distributions, energy-loudness coupling,
genre-level differences in danceability, and pairwise feature correlations.

# 2. My Role on the Team

**Primary role: Person 4 — Analysis & Interpretation Lead.**

The three core responsibilities of this role were:

1. **Writing explanations for each plot.** I authored the figure-by-figure
   descriptions in Sections 5.1–5.4 of the team paper and the full Section 5.5
   *Discussions and Interpretation*, which moves beyond describing each figure
   to explaining *why* the pattern holds (production conventions behind the
   energy-loudness coupling, cultural structure behind the bimodal tempo, the
   selection bias behind the danceability skew, and the downstream
   implications for dimensionality reduction and genre classification). I also
   wrote the paper's *Abstract* and Section 1.1 *Background and Motivation*.

2. **Identifying patterns and trends.** To make sure the claims in my prose
   were actually supported by the data and not just visual impressions from
   the plots, I wrote independent R code to surface the underlying structure
   — modal bins for the danceability spike, density peaks for the bimodal
   tempo, Pearson correlations for the energy-loudness coupling, and a
   one-way ANOVA across the top 10 genres to test whether the danceability
   differences were real. That code is in Sections 3 and 4 of this portfolio.

3. **Connecting visuals to research questions.** Each interpretation I wrote
   is tied back to one of the four research questions posed in Section 1.3.
   Section 5 of this portfolio makes that mapping explicit so it is clear how
   each figure and each statistical check answers a specific question.

The R code below is the original code I authored in service of those
responsibilities. It is not a re-run of the team's final visualization
pipeline; it is the pattern-identification and verification code I wrote so
that the explanations in Sections 5.1–5.5 of the paper could be stated with
confidence.

# 3. Data Loading and Preprocessing Review

```{r load}
library(dplyr)
library(ggplot2)

spotify_df <- read.csv("genres_v2.csv", stringsAsFactors = FALSE)

# Overall shape
dim(spotify_df)
length(unique(spotify_df$genre))
```

## 3.1 Missing-value audit

a per-column NA audit rather than relying on a single `sum(is.na())` call was used, because missingness patterns matter for which features we can interpret safely.

```{r na-audit}
na_counts <- colSums(is.na(spotify_df))
na_report <- data.frame(
  feature = names(na_counts),
  n_missing = as.integer(na_counts),
  pct_missing = round(100 * na_counts / nrow(spotify_df), 3)
) |> arrange(desc(n_missing))

na_report
```

## 3.2 Duplicate-track audit

```{r dupes}
if ("id" %in% names(spotify_df)) {
  n_total <- nrow(spotify_df)
  n_unique_ids <- length(unique(spotify_df$id))
  cat("Total rows:", n_total, "\n")
  cat("Unique track IDs:", n_unique_ids, "\n")
  cat("Duplicated ID rows:", n_total - n_unique_ids, "\n")
}
```

## 3.3 Range / sanity check on numeric features

Several Spotify features are documented as bounded to [0, 1] (danceability,
energy, valence, acousticness, speechiness, instrumentalness, liveness).an important check was ensuring the data respected those biunds before interpreting any "peaks near 0.78" or "skews above 0.75" statements in the discussion section.

```{r range-check}
bounded_features <- c("danceability", "energy", "valence",
                      "acousticness", "speechiness",
                      "instrumentalness", "liveness")

bounded_features <- bounded_features[bounded_features %in% names(spotify_df)]

range_report <- sapply(spotify_df[bounded_features], function(x) {
  c(min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    mean = round(mean(x, na.rm = TRUE), 3))
})

t(range_report)
```

## 3.4 Genre representation

The paper filters to the top 10 most represented genres for the boxplot. I
wanted to see the full genre distribution to justify that cutoff — i.e., to
confirm that a long tail of sparsely represented genres exists and would have
distorted the comparison if included.

```{r genre-counts}
genre_counts <- spotify_df |>
  count(genre, name = "n_tracks") |>
  arrange(desc(n_tracks))

genre_counts
```

# 4. Statistical Backing for the Section 5.5 Interpretations

Each subsection below corresponds to a claim I made in Section 5.5. The goal
was not to produce new visualizations (the team had those) but to compute the
underlying numbers so my prose claims were defensible.

## 4.1 Figure 1 claim — "notable spike near 0.78"

In Section 5.5 I described a pronounced peak in the danceability histogram
around 0.78. I verified this with a targeted summary rather than eyeballing
the plot.

```{r dance-peak}
dance <- spotify_df$danceability[!is.na(spotify_df$danceability)]

summary(dance)

# Where is the modal bin at ~0.01 resolution?
breaks <- seq(0, 1, by = 0.01)
bin_counts <- table(cut(dance, breaks = breaks, include.lowest = TRUE))
top_bins <- sort(bin_counts, decreasing = TRUE)[1:5]
top_bins
```

## 4.2 Figure 2 claim — energy is right-skewed

My 5.5 paragraph on Figure 2 leaned on the dataset being dominated by
high-energy tracks (values above 0.75). I checked this numerically.

```{r energy-skew}
energy <- spotify_df$energy[!is.na(spotify_df$energy)]
mean(energy)
median(energy)
mean(energy > 0.75)   # share of tracks above 0.75
mean(energy < 0.25)   # share of tracks below 0.25
```

## 4.3 Figure 3 claim — bimodal tempo distribution

The most specific interpretive claim in 5.5 is that tempo has two modes, one
near 125–130 BPM (hip-hop family) and one near 150 BPM (techno/techhouse
family). I confirmed bimodality by locating the density peaks directly from a
kernel density estimate.

```{r tempo-bimodal}
tempo <- spotify_df$tempo[!is.na(spotify_df$tempo)]
d <- density(tempo, adjust = 1)

# Local maxima on the density curve
is_peak <- c(FALSE, diff(sign(diff(d$y))) < 0, FALSE)
peaks <- data.frame(bpm = round(d$x[is_peak], 1),
                    density = round(d$y[is_peak], 5))
peaks <- peaks[order(-peaks$density), ]
head(peaks, 4)
```

## 4.4 Figure 4 claim — energy and loudness are positively related

```{r energy-loudness}
cor(spotify_df$energy, spotify_df$loudness,
    use = "complete.obs", method = "pearson")
```

## 4.5 Figure 5 claim — genre meaningfully shifts danceability

To back up the claim that genre is a meaningful predictor of danceability — as
opposed to the boxplot simply looking different by eye — I compared group means
across the top 10 genres with a one-way ANOVA.

```{r genre-anova}
top10 <- genre_counts$genre[1:10]

sub <- spotify_df |>
  filter(genre %in% top10) |>
  select(genre, danceability) |>
  filter(!is.na(danceability))

aov_fit <- aov(danceability ~ genre, data = sub)
summary(aov_fit)

# Group medians, ordered
sub |>
  group_by(genre) |>
  summarise(median_dance = round(median(danceability), 3),
            iqr = round(IQR(danceability), 3),
            n = n()) |>
  arrange(desc(median_dance))
```

## 4.6 Figure 6 claim — strongest correlations are energy-loudness (+0.60) and energy-acousticness (-0.50)

```{r corr-matrix}
numeric_feats <- c("danceability", "energy", "loudness", "speechiness",
                   "acousticness", "instrumentalness", "liveness",
                   "valence", "tempo")
numeric_feats <- numeric_feats[numeric_feats %in% names(spotify_df)]

corr_mat <- cor(spotify_df[numeric_feats], use = "complete.obs")
round(corr_mat, 2)

# Extract the two pairs I called out in the paper
round(corr_mat["energy", "loudness"], 2)
round(corr_mat["energy", "acousticness"], 2)
```

# 5. Connecting Visuals to Research Questions

A key part of my role was making sure every figure in the paper actually
serves one of the four research questions from Section 1.3. The table below
is the mapping I used while writing Sections 5.1–5.5, and the checks in
Section 4 of this portfolio line up one-to-one with these.

| Research Question | Figure(s) | Numerical Check (§4) | Interpretation Written |
|---|---|---|---|
| RQ1. How are audio features distributed across the dataset? | Fig. 1 (danceability), Fig. 2 (energy), Fig. 3 (tempo) | §4.1 modal bin; §4.2 skew share; §4.3 density peaks | §5.1 and Fig. 1–3 paragraphs in §5.5 |
| RQ2. Is there a measurable relationship between energy and loudness? | Fig. 4 (scatter with trend line) | §4.4 Pearson r | §5.2 and Fig. 4 paragraph in §5.5 |
| RQ3. Do music genres differ significantly in danceability? | Fig. 5 (boxplot, top 10 genres) | §4.5 one-way ANOVA + group medians | §5.3 and Fig. 5 paragraph in §5.5 |
| RQ4. Which audio features are most strongly correlated? | Fig. 6 (correlation heatmap) | §4.6 correlation matrix | §5.4 and Fig. 6 paragraph in §5.5 |

This mapping was what I used to keep the interpretation sections disciplined.
Every paragraph I wrote in 5.5 had to answer "which research question does
this serve?" before it was allowed in, which is also why the discussion stops
at six figure-paragraphs rather than sprawling into unrelated commentary.

# 6. Reflection and Learning Outcomes

Working on the interpretation side of the paper taught me that an EDA is only
as trustworthy as the cleaning decisions behind it. Data is often times open to interpretation and it takes a lot of patience to understand what an underlying dataset contains.

In Section 5.5 I made some fairly strong claims — "tempo splits into two
genre communities," "genre meaningfully shifts danceability" — that could have
easily been overstated. If i were to extend this work, i would work more with formal sensitivity analysis. this would show how different findings survive.



# 7. Repository and Attribution

- Team repository: `CSC_308_HamptonU_Final_Project` (GitHub, public)
- This portfolio: `Ezaria_Alexander_Portfolio.Rmd`

All R code in this portfolio was written by me. Narrative sections I authored
on the team paper are the introductions and the full Section 5.5 *Discussions and Interpretation*.
