---
author: "Makena Jelks"
date: April 16, 2026
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

## Overview of Contribution

This R Markdown documents my individual contribution to the team project
**"Exploratory Analysis of Spotify Audio Features Across Music Genres Using R."**
My primary responsibilities were:

- Locating and loading the dataset
- Performing all data cleaning and preprocessing
- Producing summary statistics and describing the dataset for the final paper
- Sharing the cleaned `spotify_df` object 

---

## 1. Data Collection

### 1.1 Source

The dataset was obtained from Kaggle, published by user *mrmorj* (2020) under the
title *"Dataset of Songs in Spotify."* The data was originally extracted from the
**Spotify Web API**, specifically the `/audio-features` endpoint, which returns
quantitative audio metrics for individual tracks.

- **Citation:** mrmorj. 2020. *Dataset of Songs in Spotify*. Kaggle.
  https://www.kaggle.com/datasets/mrmorj/dataset-of-songs-in-spotify
- **File used:** `genres_v2.csv`
- **Access method:** Downloaded from Kaggle and loaded locally via `read.csv()`

```{r load-data}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)

# Load the dataset
spotify_df_raw <- read.csv("genres_v2.csv", stringsAsFactors = FALSE)

# Quick peek at structure
cat("Dimensions (rows x cols):", nrow(spotify_df_raw), "x", ncol(spotify_df_raw), "\n")
cat("Column names:\n")
print(names(spotify_df_raw))
```

---

## 2. Data Description (Raw)

### 2.1 Basic Properties

```{r raw-description}
cat("Total tracks:", nrow(spotify_df_raw), "\n")
cat("Total features (columns):", ncol(spotify_df_raw), "\n")
cat("Unique genres:", length(unique(spotify_df_raw$genre)), "\n")
cat("Genres present:\n")
print(sort(unique(spotify_df_raw$genre)))
```

### 2.2 Missing Values (Raw Dataset)

```{r missing-raw}
missing_counts <- colSums(is.na(spotify_df_raw))
missing_df <- data.frame(
  Feature = names(missing_counts),
  Missing_Count = missing_counts,
  Missing_Pct = round(missing_counts / nrow(spotify_df_raw) * 100, 1)
)
kable(missing_df[missing_df$Missing_Count > 0, ], row.names = FALSE,
      caption = "Columns with missing values in raw dataset")
```

> **Note:** Missing values appear only in `song_name`, `Unnamed: 0`, and `title` — 
> metadata columns that are not used in our analysis. All 14 audio feature columns 
> and the `genre` label have zero missing values.

### 2.3 Duplicate Rows

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(spotify_df_raw)), "\n")
```

---

## 3. Data Cleaning

The following steps were taken to prepare a clean `spotify_df` for team use.

### Step 1 — Retain only analytical columns

We drop administrative/metadata columns (`Unnamed: 0`, `type`, `id`, `uri`,
`track_href`, `analysis_url`, `title`) that carry no analytical value. We retain
`song_name` as a label even though it has many missing values, since it is not
used in any computation.

```{r clean-step1}
# Columns to keep
keep_cols <- c("danceability", "energy", "key", "loudness", "mode",
               "speechiness", "acousticness", "instrumentalness",
               "liveness", "valence", "tempo", "duration_ms",
               "time_signature", "genre", "song_name")

spotify_df <- spotify_df_raw %>%
  select(all_of(keep_cols))

cat("Columns after selection:", ncol(spotify_df), "\n")
```

### Step 2 — Ensure correct data types

```{r clean-step2}
# Numeric audio features
num_features <- c("danceability", "energy", "loudness", "speechiness",
                  "acousticness", "instrumentalness", "liveness",
                  "valence", "tempo", "duration_ms")

spotify_df <- spotify_df %>%
  mutate(across(all_of(num_features), as.numeric),
         key            = as.integer(key),
         mode           = as.factor(mode),          # 0 = minor, 1 = major
         time_signature = as.integer(time_signature),
         genre          = as.factor(genre))

cat("Data types after conversion:\n")
print(sapply(spotify_df, class))
```

### Step 3 — Remove rows with NA in core audio features

```{r clean-step3}
before <- nrow(spotify_df)
spotify_df <- spotify_df %>%
  filter(complete.cases(select(., all_of(num_features))))
after <- nrow(spotify_df)

cat("Rows before:", before, "\n")
cat("Rows after removing NA in audio features:", after, "\n")
cat("Rows removed:", before - after, "\n")
```

### Step 4 — Remove physical outliers

Loudness values above 0 dB are physically impossible for normalized tracks;
duration of 0 ms indicates a corrupt record.

```{r clean-step4}
before <- nrow(spotify_df)
spotify_df <- spotify_df %>%
  filter(loudness <= 0,
         duration_ms > 0,
         tempo > 0)
after  <- nrow(spotify_df)

cat("Rows removed by outlier filter:", before - after, "\n")
cat("Final cleaned dataset size:", nrow(spotify_df), "tracks\n")
```

### Step 5 — Verify final cleanliness

```{r clean-step5}
cat("Missing values in cleaned spotify_df:\n")
print(colSums(is.na(spotify_df[, num_features])))
cat("\nDuplicate rows:", sum(duplicated(spotify_df)), "\n")
```

---

## 4. Cleaned Dataset — Summary Statistics

```{r summary-stats}
summary_tbl <- spotify_df %>%
  select(all_of(num_features)) %>%
  summarise(across(everything(),
    list(
      mean   = ~round(mean(.),   3),
      median = ~round(median(.), 3),
      sd     = ~round(sd(.),     3),
      min    = ~round(min(.),    3),
      max    = ~round(max(.),    3)
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  pivot_longer(everything(),
               names_to  = c("Feature", "Stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = Stat, values_from = value)

kable(summary_tbl, caption = "Descriptive statistics for numeric audio features")
```

### Genre Distribution

```{r genre-dist}
genre_counts <- spotify_df %>%
  count(genre, sort = TRUE) %>%
  rename(Count = n) %>%
  mutate(Percent = round(Count / sum(Count) * 100, 1))

kable(genre_counts, caption = "Track counts per genre")

ggplot(genre_counts, aes(x = reorder(genre, Count), y = Count, fill = genre)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Track Count per Genre",
       x = "Genre", y = "Number of Tracks") +
  theme_minimal()
```

---

## 5. Data Dictionary

```{r data-dict}
dict <- data.frame(
  Feature = c("danceability","energy","key","loudness","mode",
              "speechiness","acousticness","instrumentalness","liveness",
              "valence","tempo","duration_ms","time_signature","genre","song_name"),
  Type = c("Numeric (0–1)","Numeric (0–1)","Integer (0–11)","Numeric (dB)",
           "Factor (0/1)","Numeric (0–1)","Numeric (0–1)","Numeric (0–1)",
           "Numeric (0–1)","Numeric (0–1)","Numeric (BPM)","Integer (ms)",
           "Integer","Factor","Character"),
  Description = c(
    "Suitability of a track for dancing; 1.0 = most danceable",
    "Intensity and activity; 1.0 = most energetic",
    "Musical key (0 = C, 1 = C#, …, 11 = B)",
    "Overall loudness in decibels; typically –60 to 0 dB",
    "Modality: 1 = major, 0 = minor",
    "Presence of spoken words; >0.66 = likely spoken word",
    "Confidence the track is acoustic; 1.0 = acoustic",
    "Probability of no vocals; >0.5 = likely instrumental",
    "Probability of a live audience; >0.8 = likely live recording",
    "Musical positiveness; 1.0 = happy/euphoric, 0.0 = sad/angry",
    "Estimated tempo in beats per minute",
    "Track duration in milliseconds",
    "Estimated time signature (beats per bar)",
    "Spotify genre label (15 unique values)",
    "Track title from Spotify (partially missing — not used in analysis)"
  )
)
kable(dict, caption = "Data Dictionary for spotify_df")
```

---

## 6. Sharing the Cleaned Data

The cleaned `spotify_df` object is saved as an `.RData` file so all team
members can load it directly without repeating the cleaning steps.

```{r save-clean, eval=FALSE}
# Run this once to produce the shared file
save(spotify_df, file = "spotify_df_clean.RData")

# Team members load it with:
# load("spotify_df_clean.RData")
```

---

## 7. Analytical Decisions & Justifications

| Decision | Justification |
|---|---|
| Dropped metadata columns (`id`, `uri`, etc.) | These are API identifiers with no analytical meaning |
| Kept `song_name` despite NAs | Not used in computation; useful for spot-checking results |
| Converted `mode` to factor | It is a binary categorical variable, not a continuous quantity |
| Filtered `loudness > 0` | Values above 0 dB violate Spotify's normalization; only 2 such rows existed |
| Used `genres_v2.csv` over `genres.csv` | Version 2 contains the genre label column required for all genre-level analysis |

---

## 8. Session Info

```{r session}
sessionInfo()
```
